pdp: An R Package for Constructing Partial Dependence Plots

### 2016-12-11 manuscript/reviews/response.Rmd

---
title: "Response to Comments"
output:
  html_document: default
  html_notebook: default
  pdf_document: default
  word_document: default
---

A few additional changes have been made to the package since the original submission. The Boston housing data no longer has to be imported from the `mlbench` package; it is loaded automatically with `pdp`. `pdp` now interfaces with the `caret` package. This is breifely discussed on page X and has been incorporated into the last example.

A few of the comments were related to the plotting behavior of `plotPartial`. I added a paragraph on page X that should help clarify many of these issues.

### 2016-12-11 manuscript/greenwell.tex
tweaks based on review comments

The \pkg{pdp} package is useful for constructing PDPs for many classes of fitted models in R. PDPs are especially useful for visualizing the relationships discovered by complex machine learning algorithms such as a random forest. The latest stable release is available from \href{https://cran.r-project.org/package=pdp}{CRAN}:
\begin{example}
install.packages("pdp")
\end{example}
The development version is located on GitHub: \url{https://github.com/bgreenwell/pdp}. Bug reports and suggestions are appreciated and should be submitted to \url{https://github.com/bgreenwell/pdp/issues}. Currently, only two functions are exported by \pkg{pdp}:
\begin{itemize}
  \item \code{partial}
  \item \code{plotPartial}
\end{itemize}

@@

\strong{Note:} as mentioned above, \pkg{pdp} relies on \pkg{lattice} for its graphics. \pkg{lattice} itself is built on top of \CRANpkg{grid} \citep{grid-pkg}. \pkg{grid} graphics behave a little differently than traditional R graphics and two points are worth making:
\begin{enumerate}
  \item \pkg{lattice} functions create a \code{"trellis"} object, but does not display it; the \code{print} method produces the actual display. The result is automatically printed when using these functions in the command line, but in \code{source} or inside a function, an explicit \code{print} statement is required to display the resulting graph.
  \item Setting graphical parameters via \code{par} typically has no effect on \pkg{lattice} plots. Instead, \pkg{lattice} provides its own \code{trellis.par.set} function for modifying graphical parameters.
\end{enumerate}
A consequence of second point is that the \code{par} function can not be used to control the layout of multiple \pkg{lattice} (and hence \pkg{pdp}) plots. Simple solutions are available in packages \CRANpkg{latticeExtra} and \CRANpkg{gridExtra}. For convenience, \pkg{pdp} imports the \code{grid.arrange} function from \pkg{gridExtra} which makes it easy to display multiple \pkg{pdp} plots on a single page.

@@

Constructing PDPs can be quite computationally expensive. Several strategies are available to ease the computational burden in larger problems. For example, there is no need to compute partial dependence of \code{cmedv} using each unique value of \code{rm} in the training data (which would require $k = 446$ passes over the data!). We could get very reasonable results using a reduced number of points. Current options are to use a grid of equally spaced values in the range of the variable of interest; the number of points can be controlled using the \code{grid.resolution} option in the call to \code{partial}. Alternatively, a user-specified grid of values (e.g., containing specific quantiles of interest) can be supplied through the \code{pred.grid} argument. To demonstrate, the following snippet of code computes the partial dependence of \code{cmedv} on \code{rm} using each option; the results are displayed in Figure~\ref{fig:partial_manual}.
\begin{example}
grid.arrange(
  # Figure 5 (left)
  partial(boston.rf, "rm", plot = TRUE),

  # Figure 5 (middle)
  partial(boston.rf, "rm", grid.resolution = 30, plot = TRUE),

  # Figure 5 (right)
  partial(boston.rf, "rm", pred.grid = data.frame(rm = 3:9), plot = TRUE),

  # One column for each PDP
  ncol = 3
)
\end{example}

@@

Setting up a parallel backend is rather straightforward. To demonstrate, the following snippet of code sets up the \code{partial} function to run in parallel on Unix-like systems\footnote{This example will not run on Windows.} using the \pkg{doParallel} package.
\begin{example}
library(doParallel)  # load parallel backend
registerDoParallel(cores = 4)  # use 4 cores
\end{example}
Now, to run \code{partial} in parallel, all we have to do is invoke the \code{parallel = TRUE} option and the rest is taken care of by the internal call to \pkg{plyr} and the parallel backend we loaded. This is illustrated in the code chunk below which obtains the partial dependence of \code{ozone} on \code{wind}, \code{temp}, and \code{dpg} in parallel. The result is displayed in Figure~\ref{fig:partial_par}.
\begin{example}
partial(ozone.mars, pred.var = c("wind", "temp", "dpg"), plot = TRUE,
        chull = TRUE, parallel = TRUE)
\end{example}

@@

library(ggplot2)
ggplot(pd, aes(x = Petal.Width, y = Petal.Length, z = y, fill = y)) +
  geom_tile() +
  geom_contour(color = "white", alpha = 0.5) +
  scale_fill_distiller(name = "Average\nlogit", palette = "Spectral") +
  theme_bw() +
  facet_grid(~ Species)
\end{example}

@@

The code chunk below uses \pkg{caret} to train an XGBoost model of a grid of tuning values using 5-fold cross-validation.
\begin{example}
# Optimize tuning parameters using 5-fold cross-validation
library(caret)

@@

# Train an XGBoost model
set.seed(202)
boston.xgb.tune <- train(x = data.matrix(subset(boston, select = -cmedv)),
                         y = boston$cmedv,
                         method = "xgbTree",
                         metric = "RMSE",
                         trControl = ctrl,
                         tuneGrid = xgb.grid)
\end{example}
The final model achieves a cross-validated RMSE and $R^2$ of 2.819 and 0.905, respectively. The following snippet of code computes the partial dependence of \code{cmedv} on both \code{rm} and \code{lstat}, individually and together. The results are displayed in Figure~\ref{fig:boston_xgb}.
\begin{example}
grid.arrange(
  # Figure 8 (left)
  partial(boston.xgb.tune, pred.var = "rm", plot = T, rug = T),

  # Figure 8 (middle)
  partial(boston.xgb.tune, pred.var = "lstat", plot = T, rug = T),

  # Figure 8 (right)
  partial(boston.xgb.tune, pred.var = c("lstat", "rm"), plot = T, chull = T),

  # One column for each PDP
  ncol = 3
)
\end{example}

@@

Since \code{train} always stores a copy of the training data as part of the \code{"train"} object in a component called \code{trainingData} (e.g., \code{boston.xgb.tune\$trainingData}) there is no need to supply the training data in the call to \code{partial}. However, this is not the case when using the \pkg{xgboost} package directly. For example, the following code fits the same model using the \code{xgboost} function.
\begin{example}
set.seed(102)  # for reproducibility
boston.xgb <- xgboost(data = data.matrix(subset(boston, select = -cmedv)),
                      label = boston$cmedv, objective = "reg:linear",
                      nrounds = 2000, max_depth = 3, eta = 0.01)
\end{example}
To use \code{partial} with the \code{"xgb.Booster"} class, we need to supply the original training data (minus the response) in the call to \code{partial}. The following snippet of code computes the partial dependence of \code{cmedv} on \code{rm} (plot not shown). \strong{Note:} while \code{xgboost} requires the training data to be an object of class \code{"matrix"}, \code{"dgCMatrix"}, or \code{"xgb.DMatrix"}, \code{partial} requires a \code{"data.frame"}.
\begin{example}
partial(boston.xgb, pred.var = "rm", plot = TRUE, rug = TRUE,
        train = subset(boston, select = -cmedv))
\end{example}

% \begin{example}
% pred.list <- list("lstat", "rm", c("lstat", "rm"))
% for (x in pred.list) {
%   partial(boston.xgb, pred.var = x, plot = TRUE, smooth = TRUE, rug = TRUE,
%           chull = TRUE, train = subset(boston, select = -cmedv))
% }
% \end{example}

>>>>

The \pkg{pdp} package is useful for constructing PDPs for many classes of fitted models in R. PDPs are especially useful for visualizing the relationships discovered by complex machine learning algorithms such as a random forest. The latest stable release is available from \href{https://cran.r-project.org/package=pdp}{CRAN}. The development version is located on GitHub: \url{https://github.com/bgreenwell/pdp}. Bug reports and suggestions are appreciated and should be submitted to \url{https://github.com/bgreenwell/pdp/issues}. Currently, only two functions are exported by \pkg{pdp}:
\begin{itemize}
  \item \code{partial}
  \item \code{plotPartial}
\end{itemize}

@@

\strong{Note:} as mentioned above, \pkg{pdp} relies on \pkg{lattice} for its graphics. \pkg{lattice} itself is built on top of \CRANpkg{grid} \citep{grid-pkg}. \pkg{grid} graphics behave a little differently than traditional R graphics and two points are worth making (see \code{?lattice} for more details):
\begin{enumerate}
  \item \pkg{lattice} functions return a \code{"trellis"} object, but do not display it; the \code{print} method produces the actual display. However, due to R's automatic printing rule, the result is automatically printed when using these functions in the command line. If \code{plotPartial} is called inside of \code{source} or inside a loop (e.g., \code{for} or \code{while}), an explicit \code{print} statement is required to display the resulting graph.
  \item Setting graphical parameters via \code{par} typically has no effect on \pkg{lattice} plots. Instead, \pkg{lattice} provides its own \code{trellis.par.set} function for modifying graphical parameters.
\end{enumerate}
A consequence of second point is that the \code{par} function cannot be used to control the layout of multiple \pkg{lattice} (and hence \pkg{pdp}) plots. Simple solutions are available in packages \CRANpkg{latticeExtra} and \CRANpkg{gridExtra}. For convenience, \pkg{pdp} imports the \code{grid.arrange} function from \pkg{gridExtra} which makes it easy to display multiple \pkg{pdp} plots on a single page.

@@

Constructing PDPs can be quite computationally expensive. Several strategies are available to ease the computational burden in larger problems. For example, there is no need to compute partial dependence of \code{cmedv} using each unique value of \code{rm} in the training data (which would require $k = 446$ passes over the data!). We could get very reasonable results using a reduced number of points. Current options are to use a grid of equally spaced values in the range of the variable of interest; the number of points can be controlled using the \code{grid.resolution} option in the call to \code{partial}. Alternatively, a user-specified grid of values (e.g., containing specific quantiles of interest) can be supplied through the \code{pred.grid} argument. To demonstrate, the following snippet of code computes the partial dependence of \code{cmedv} on \code{rm} using each option, then uses \code{grid.arrange} to display all three PDPs on the same page; the results are displayed in Figure~\ref{fig:partial_manual}.
\begin{example}
p1 <- partial(boston.rf, "rm", plot = TRUE)
p2 <- partial(boston.rf, "rm", grid.resolution = 30, plot = TRUE)
p3 <- partial(boston.rf, "rm", pred.grid = data.frame(rm = 3:9), plot = TRUE)
grid.arrange(
  p1,  # Figure 5 (left)
  p2,  # Figure 5 (middle)
  p3,  # Figure 5 (right)
  ncol = 3
)
\end{example}

@@

Setting up a parallel backend is rather straightforward. To demonstrate, the following snippet of code sets up the \code{partial} function to run in parallel on both Windows and Unix-like systems using the \pkg{doParallel} package.
\begin{example}
library(doParallel)  # load parallel backend
cl <- makeCluster(4)  # use 4 workers
registerDoParallel(cl)
\end{example}
Now, to run \code{partial} in parallel, all we have to do is invoke the \code{parallel = TRUE} and \code{paropts} options and the rest is taken care of by the internal call to \pkg{plyr} and the parallel backend we loaded\footnote{Notice we have to pass the names of external packages that the tasks depend on via the \code{paropts} argument; in this case, \code{"earth"}. See \code{?plyr::adply} for details.}. This is illustrated in the code chunk below which obtains the partial dependence of \code{ozone} on \code{wind}, \code{temp}, and \code{dpg} in parallel. The result is displayed in Figure~\ref{fig:partial_par}. \strong{Note:} it is considered good practice to shut down the workers by calling \code{stopCluster} when finished.
\begin{example}
partial(ozone.mars, pred.var = c("wind", "temp", "dpg"), plot = TRUE,
        chull = TRUE, parallel = TRUE, paropts = list(.packages = "earth"))  # Figure 6
stopCluster(cl)  # good practice
\end{example}

@@

# Figure 7
library(ggplot2)
ggplot(pd, aes(x = Petal.Width, y = Petal.Length, z = y, fill = y)) +
  geom_tile() +
  geom_contour(color = "white", alpha = 0.5) +
  scale_fill_distiller(name = "Logit", palette = "Spectral") +
  theme_bw() +
  facet_grid(~ Species)
\end{example}

@@

The code chunk below uses \pkg{caret} to train an XGBoost model over a grid of tuning values using 5-fold cross-validation.
\begin{example}
# Optimize tuning parameters using 5-fold cross-validation
library(caret)

@@

# Train an XGBoost model
set.seed(202)
boston.xgb <- train(x = data.matrix(subset(boston, select = -cmedv)),
                    y = boston$cmedv,
                    method = "xgbTree",
                    metric = "RMSE",
                    trControl = ctrl,
                    tuneGrid = xgb.grid)
\end{example}
The model with the smallest root-mean-squared error (RMSE) had \code{nrounds = 2000}, \code{max\_depth = 3}, \code{eta = 0.01}, and achieved a cross-validated RMSE and $R^2$ of 2.819 and 0.905, respectively. The following snippet of code computes the partial dependence of \code{cmedv} on both \code{rm} and \code{lstat}, individually and together, and then uses \code{grid.arrange} to display al three PDPs on the same page. The results are displayed in Figure~\ref{fig:boston_xgb}.
\begin{example}
p1 <- partial(boston.xgb, pred.var = "rm", plot = T, rug = T)
p2 <- partial(boston.xgb, pred.var = "lstat", plot = T, rug = T)
p3 <- partial(boston.xgb, pred.var = c("lstat", "rm"), plot = T, chull = T),
grid.arrange(
  p1,  # Figure 8 (left)
  p2,  # Figure 8 (middle)
  p3,  # Figure 8 (right)
  ncol = 3
)
\end{example}

@@

Since \code{train} always stores a copy of the training data as part of the \code{"train"} object in a component called \code{trainingData} (e.g., \code{boston.xgb.tune\$trainingData}) there is no need to supply the training data in the call to \code{partial}. However, this is not the case when using the \pkg{xgboost} package directly. For example, the following code fits the same model using the \code{xgboost} function.
\begin{example}
set.seed(203)  # for reproducibility
boston.xgb <- xgboost(data = data.matrix(subset(boston, select = -cmedv)),
                      label = boston$cmedv, objective = "reg:linear",
                      nrounds = 2000, max_depth = 3, eta = 0.01)
\end{example}
To use \code{partial} with the \code{"xgb.Booster"} class, we need to supply the original training data (minus the response) in the call to \code{partial}. The following snippet of code computes the partial dependence of \code{cmedv} on \code{rm} (plot not shown). \strong{Note:} while \code{xgboost} requires the training data to be an object of class \code{"matrix"}, \code{"dgCMatrix"}, or \code{"xgb.DMatrix"}, \code{partial} requires a \code{"data.frame"}.
\begin{example}
partial(boston.xgb, pred.var = "rm", plot = TRUE, rug = TRUE,
        train = subset(boston, select = -cmedv))
\end{example}
